---
title: "Getting Started with ggBoostedTrees"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with ggBoostedTrees}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## What is this package for?

If you have grown a random forest with `randomForestSRC`, you have probably
reached for `ggRandomForests` to look at it: an error curve to see whether the
forest has enough trees, an importance plot, a partial dependence plot or two.
ggBoostedTrees gives you the same set of figures for a boosted tree model of a
longitudinal outcome, fit with `boostmtree`.

Boosting adds a few questions a forest never raises. A boosted model is grown
one small step at a time, so you need to know where to stop it. `boostmtree`
also estimates a within-subject correlation and a smoothing penalty as it goes,
and those should settle before boosting stops. And because the response is
measured repeatedly on the same subjects, the figure you most want is the one
that puts a subject's fitted curve next to their observed values.

This vignette is for the biostatistician who already knows R and ggplot2 and
has a `boostmtree` fit (or is about to make one). It walks one small
simulated example from the fit to every figure the package draws, in the
order you would look at them.

## How is a figure built?

Every figure comes from two steps. An extractor, one of the `gg_boost_*()`
functions, pulls a tidy data frame out of the fitted model. Then `autoplot()`
(or its alias `plot()`) draws that data frame and hands you back a bare
ggplot.

The split is deliberate. The data frame has a documented column contract, so
you can read the numbers behind a figure, test them, or reshape them before
anything is drawn. And because the renderer reads only that data frame, never
the fit, a second modelling backend needs new extractor methods and nothing
else. That is how `BoostMLR` fits came to be supported.

## A small fit

We simulate 25 subjects with up to four visits each, using `simLong()` from
`boostmtree`. Model 1 builds the response from `x1`, `x3` and `x4`, plus an
interaction between time and `x2`, so we know what a sensible fit ought to
find. Fifty boosting iterations is enough to make the point and fits in about
a second.

```{r fit}
library(ggBoostedTrees)
library(ggplot2)

set.seed(7)
sim <- boostmtree::simLong(n = 25, n.time = 4, model = 1)$data.list

fit <- boostmtree::boostmtree(
  x = sim$features, tm = sim$time, id = sim$id, y = sim$y,
  M = 50, cv.flag = TRUE, verbose = FALSE,
  control = boostmtree::boostmtree.control(seed = 7)
)
```

`cv.flag = TRUE` matters here. `boostmtree` records the error path and the
cross-validated optimal iteration (`m.opt`) only when cross-validation ran. A
fit grown without it has no error path to draw, and `gg_boost_error()` will
tell you so rather than draw something misleading.

## Did it converge?

The error path is the first thing worth looking at, so a fitted model goes
straight to it:

```{r error-shortcut}
autoplot(fit)
```

The dashed line marks the optimal iteration. Error that falls and then
flattens well before the last iteration says `M` was large enough. If the
curve is still falling at the right edge, grow more iterations.

The shortcut hides the first step. Written out, it is the extractor and then
the renderer, and the extractor's result is an ordinary data frame:

```{r error-data}
gg_dta <- gg_boost_error(fit)
head(gg_dta, 3)
```

The `optimal` column flags the row for `m.opt`. Everything else in the
package follows this pattern.

## Did the variance structure settle?

`boostmtree` re-estimates three parameters as it boosts: `rho` (the
within-subject correlation), `phi` (the variance component) and `lambda` (the
P-spline smoothing parameter for the time interaction). `gg_boost_path()`
tracks all three by iteration:

```{r path}
autoplot(gg_boost_path(fit))
```

Each parameter gets its own panel on a free y scale, because the three can
sit orders of magnitude apart. What you want to see is each path levelling
off. A parameter still drifting at the last iteration means the variance
model had not caught up with the mean model when boosting stopped.

## Does it track individual subjects?

This is the figure longitudinal boosting exists for. A model can match the
population mean well and still miss most individual subjects.
`gg_boost_trajectory()` puts each subject's fitted curve (lines) over their
observed values (points):

```{r trajectory}
autoplot(gg_boost_trajectory(fit))
```

With a few hundred subjects the plot turns to spaghetti. `autoplot()` draws at
most `n_max` subjects (100 by default) and scales the transparency to the
number drawn. Use `subset` to name the subjects you want to look at.

## Does it match the data over follow-up?

The trajectory plot shows subjects one at a time. The calibration figure
summarizes all of them at once: observations are grouped into time bins with
roughly equal counts, and each bin shows the mean observed value with a 95%
interval next to the mean fitted value at the same observations.

```{r calibration}
pred <- predict(fit, x = fit$x)
autoplot(gg_boost_calibration(fit, pred = pred))
```

Passing `pred` adds the cohort's mean predicted curve (the solid line). The
interval treats observations as independent, which repeated measurements are
not, so it is too narrow. Read it as a visual guide, not a test. The extracted
data frame also carries `n_subject` per bin, which is where to look when a
late bin rests on a handful of subjects.

## Which covariates matter, and how?

Variable importance comes from `boostmtree` itself. `gg_boost_vimp()` takes
the output of `vimp.boostmtree()` and splits importance into the covariate's
main effect and its interaction with time:

```{r vimp}
set.seed(7)
vimp <- boostmtree::vimp.boostmtree(fit)
autoplot(gg_boost_vimp(vimp))
```

Don't read too much into importance from 25 subjects. The values are noisy at
this size. A value below zero means permuting that covariate did not make the
error worse, so the model is not leaning on it. Here only the main effect of
`x1` clears zero, and the simulation did give `x1` the largest coefficient.

To see the shape of an effect rather than its size, ask `boostmtree` for a
partial effect and hand the result to `gg_boost_effect()`:

```{r effect}
pp <- boostmtree::partial.plot(
  fit, x.var.names = "x1", output = "data", verbose = FALSE
)
autoplot(gg_boost_effect(pp))
```

Each curve is the partial effect of `x1` at one time point, coloured by time.
The response rises with `x1`, in steps, because the base learners are trees.
The vertical gap between curves is the time trend. Curves that stay roughly
parallel, as these do, say the effect of `x1` does not change much with time;
a covariate with a strong time interaction draws curves that change slope
from one time to the next. `gg_boost_effect()` accepts `marginal.plot()`
output in the same way.

## Finishing a figure

`autoplot()` returns a bare ggplot, so you finish it with `+` the way you
would any other:

```{r finish}
autoplot(gg_dta) +
  theme_bw() +
  labs(title = "Cross-validated error by iteration")
```

Swap in the `hvtiPlotR` theme here for a figure in house style.

## What about BoostMLR?

All six figures also accept a `BoostMLR` fit. Partial effects go through
`boostmlr_partial()`, since `BoostMLR`'s own partial-effect output arrives
without covariate or response names. The support is partial on purpose:
marginal effects and joint importance are not available for `BoostMLR`, and
the extractors stop with an error rather than guess.

## Where to go next

The [function reference](https://ehrlinger.github.io/ggBoostedTrees/reference/)
is arranged in the same order as this vignette's workflow:

- **Extract** holds the `gg_boost_*()` extractors and `boostmlr_partial()`,
  each documenting the columns of the data frame it returns.
- **Render** holds the `autoplot()` method for each extracted class, with the
  arguments that control the drawing (`subset` and `n_max` for trajectories,
  for example).
- **Shortcut** holds `autoplot.boostmtree()`, the jump from a fit to its error
  path.

For random forests, the same idiom is in
[ggRandomForests](https://github.com/ehrlinger/ggRandomForests).
